---
title: "Testing External Pointers"
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

```{r include = FALSE}
library(xptrpkgs)
```

This package comes with two functions, `make` and `release`, which are C++ functions. `make(i, j)` returns an _external pointer_ to a `arma::Mat<short>` object with `i` rows and `j` columns and `release` frees the memory of an external pointer. For example:

```{r}
x <- make(1e3, 1e6)
```

which takes up approximately 1.9Gb on my machine (using `free -m` on linux). We can then free the memory of `x` by

```{r}
release(x)
```

We can also register a finalizer on the `R` side:

```{r}
f <- function(x) release(x)
reg.finalizer(x, f)
```

This will ensure, that `x` is garbage collected when `gc()` is run. However, `R` does not "see" the actual memory footprint of `x`:

```{r}
format(object.size(x), units = "b", standard = "auto", digits = 1L)

# cleaning
rm(x)
gc()
```

Thus, creating external pointers does not set aside enough memory if we hope these to be garbage collected at some point.

However, for very large objects, there is a huge benefit staying on the `C++` side using external pointers, since similar objects in `R` takes up *much more* memory. The equvivalent to `x` takes:

```{r}
y <- matrix(1L, nrow = 1000L, ncol = 1e6)
format(object.size(y), units = "Gb", standard = "auto", digits = 1L)

# cleaning
rm(y)
gc()
```

Of course this is also a result of representing the matrix with `short` integers on C++; but that is the exact benefit we hope to exploit.

Is there, by any means, a way to achive the above? That is, can we design us out of the problem. If so, I have no idea where to start and I welcome any advise.
